---
title: Publications
---

<!-- After the .bib file is updated, knit this Rmd to update publications.html, then commit and push both the .bib and .html -->

(Under construction)

## Recent

Mukhopadhyay, P., Ye, J., Anderson, K. M., Roychoudhury, S., Rubin, E. H., Halabi, S., & Chappell, R. J. (2022). Log-Rank Test vs MaxCombo and Difference in Restricted Mean Survival Time Tests for Comparing Survival Under Nonproportional Hazards in Immuno-oncology Trials: A Systematic Review and Meta-analysis. JAMA oncology.

Anderson, K. M., Guo, Z., Zhao, J., & Sun, L. Z. (2022). A unified framework for weighted parametric group sequential design. Biometrical Journal.

Chen, T. Y., Zhao, J., Sun, L., & Anderson, K. M. (2021). Multiplicity for a group sequential trial with biomarker subpopulations. Contemporary Clinical Trials, 101, 106249.

Bautista, O., & Anderson, K. (2021). Sample Size Estimation and Power Analysis: Time to Event Data. In Handbook of Statistical Methods for Randomized Controlled Trials (pp. 275-300). Chapman and Hall/CRC.

Roychoudhury, S., Anderson, K. M., Ye, J., & Mukhopadhyay, P. (2021). Robust design and analysis of clinical trials with nonproportional hazards: a straw man guidance from a cross-pharma working group. Statistics in Biopharmaceutical Research, 1-15.

Lin, R. S., Lin, J., Roychoudhury, S., Anderson, K. M., Hu, T., Huang, B., ... & Working, C. P. N. P. H. (2020). Rejoinder to letter to the editor “The hazards of period specific and weighted hazard ratios”. Statistics in Biopharmaceutical Research, 12(4), 520.

Lin, R. S., Lin, J., Roychoudhury, S., Anderson, K. M., Hu, T., Huang, B., ... & Cross-Pharma Non-Proportional Hazards Working Group. (2020). Alternative analysis methods for time to event endpoints under nonproportional hazards: a comparative analysis. Statistics in Biopharmaceutical Research, 12(2), 187-198.

Tian, L., Jin, H., Uno, H., Lu, Y., Huang, B., Anderson, K. M., & Wei, L. J. (2020). On the empirical choice of the time window for restricted mean survival time. Biometrics, 76(4), 1157-1166.

León, L. F., Lin, R., & Anderson, K. M. (2020). On weighted log-rank combination tests and companion Cox model estimators. Statistics in Biosciences, 12(2), 225-245.

Turner, D. C., Kondic, A. G., Anderson, K. M., Robinson, A. G., Garon, E. B., Riess, J. W., ... & Stone, J. A. (2018). Pembrolizumab Exposure–Response Assessments Challenged by Association of Cancer Cachexia and Catabolic ClearancePembrolizumab PK/PD Confounding and Cancer Cachexia. Clinical Cancer Research, 24(23), 5841-5849.

Joseph, Richard W., et al. "Baseline Tumor Size Is an Independent Prognostic Factor for Overall Survival in Patients with Melanoma Treated with Pembrolizumab: Impact of Baseline Tumor Size on Outcomes in Melanoma." *Clinical Cancer Research* *24.20* (2018): 4960-4967.

Chen, Cong, et al. "A 2-in-1 adaptive phase 2/3 design for expedited oncology drug development." Contemporary clinical trials 64 (2018): 238-242.

```{r, include=FALSE}
df <- bib2df::bib2df("keaven.bib") |> suppressWarnings()
df <- df[order(df$YEAR, decreasing = TRUE), ]

# Copy the assets into a temporary directory to render
create_tempdir <- function(name) {
  path <- file.path(tempdir(), name)
  if (dir.exists(path)) unlink(path, recursive = TRUE)
  dir.create(path)
  invisible(path)
}

path_dir <- create_tempdir("pub")
path_rmd <- file.path(path_dir, "pub.Rmd")
path_md <- file.path(path_dir, "pub.md")
file.copy(from = "keaven.bib", to = path_dir, overwrite = TRUE) |> invisible()
file.copy(from = "apa-numeric-superscript-brackets.csl", to = path_dir, overwrite = TRUE) |> invisible()

c(
  "---",
  "output: md_document",
  "bibliography: keaven.bib",
  "csl: apa-numeric-superscript-brackets.csl",
  "---",
  "",
  paste0("@", df$BIBTEXKEY)
) |>
  writeLines(con = path_rmd)

withr::with_dir(
  path_dir,
  rmarkdown::render(path_rmd, output_file = path_md)
)

# Remove the reference anchors
md_text <- path_md |>
  readLines() |>
  gsub("^<sup>.*</sup>", "", x = _) |>
  gsub("^\\d+\\\\\\.\\s", "- ", x = _)

# Remove the remaining blank lines at top
md_text <- md_text[-(1L:(which(md_text != "")[1] - 2L))]

# Tag the first article of that year with year headings
idx_year <- which(md_text == "")
v <- cumsum(rle(df$YEAR)$lengths) + 1L
idx_first_year <- c(1L, v[-length(v)])
md_text[idx_year[idx_first_year]] <- paste0("## ", df$YEAR[idx_first_year])

# Add blank line around year headings
# <https://stackoverflow.com/a/18951302>
insert_at <- function(x, pos, lst) {
  stopifnot(length(lst) == length(pos))
  res <- vector("list", 2 * length(pos) + 1)
  res[c(TRUE, FALSE)] <- split(x, cumsum(seq_along(x) %in% (pos + 1)))
  res[c(FALSE, TRUE)] <- lst
  unlist(res)
}

md_text <- md_text |>
  insert_at(
    c(idx_year[idx_first_year] - 1L, idx_year[idx_first_year])[-1],
    as.list(rep("", length(idx_first_year) * 2 - 1))
  )
```

```{r, echo=FALSE, results="asis"}
cat(paste0(md_text, collapse = "\n"))
```
