---
title: "Smoothing hazard rates and survival curves with bshazard"
author: Keaven Anderson, Nan Xiao
date: "2024-09-12"
slug: "bshazard"
tags:
  - survival analysis
  - hazard function
  - smoothing splines
  - Kaplan-Meier curves
  - time-to-event
  - bshazard
  - simtrial
description: >
    Learn how to smooth hazard rates and survival curves using the R package
    bshazard. This post provides practical examples, code snippets,
    and visualizations to identify crossing points in hazard rates,
    compare smoothed estimates with Kaplan-Meier curves, and validate
    model fits in survival analysis.
bibliography: "bshazard.bib"
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  out.width = "100%",
  fig.align = "center",
  fig.width = 7 / 1.2,
  fig.height = 5 / 1.2,
  dpi = 300
)
```

<p></p>

![National Aquarium Denmark, Den Blå Planet. Photo by [Sandro Katalina](https://unsplash.com/photos/sRaRlaFQolw).](images/sandro-katalina-sRaRlaFQolw-unsplash.jpg)

## Introduction

The R package [bshazard](https://cran.r-project.org/package=bshazard)
[@rebora2014bshazard] has been available on CRAN since 2014.
It might be useful where there is interest in where smoothed hazard rates cross.
When there is excess in one arm early and excess in the other arm late,
the time at which (smoothed) hazard rates cross can come well before the
time when Kaplan-Meier survival curves cross.
The time at which hazard rates cross is probably a more important demarcation
for the period of early excess risk than the time at which survival curves cross.

As noted by @rebora2014bshazard, another alternative is the
[flexsurv](https://cran.r-project.org/package=flexsurv) package
for comparison [@jackson2016flexsurv].
The bshazard package has no unit testing included, but we have found the
model checking procedure described below adequate to justify the use of the package.
This sort of model checking is likely to be recommended regardless of how
hazard rates are estimated.

```{r, message=FALSE, warning=FALSE}
library(bshazard)
library(simtrial)
library(dplyr)
library(survival)
library(ggsurvfit)
```

```{r, message=FALSE, warning=FALSE}
# Crossing survival curves example from simtrial
data(ex6_crossing)
```

```{r}
# Set theme and color scales globally
ggplot2::theme_set(cowplot::theme_cowplot())

# Observable 10 color palette
pal <- c("#EFB118", "#4269D0", "#9498A0")
pal_ctrl <- pal[1]
pal_expt <- pal[2]
pal_gray <- pal[3]
scale_color <- function(...) scale_color_manual(values = pal, ...)
scale_fill <- function(...) scale_fill_manual(values = pal, ...)
options(ggplot2.discrete.colour = scale_color)
options(ggplot2.discrete.fill = scale_fill)
```

## Example

For the following survival curves based on reverse-engineered survival data
from the [simtrial](https://merck.github.io/simtrial/) package,
the estimated Kaplan-Meier curves cross between 10 and 11 months.
However, we see that the widest gap between the survival curves may have been
around 3 months, with the curves converging thereafter.
We will evaluate this further using B-spline smoothed hazard rate estimates
in the next section.

```{r}
#| kaplan-meier-survival-curves,
#| fig.height = 5.3 / 1.2,
#| fig.cap="Kaplan-Meier survival curves for Control and Treatment groups, with risk tables displaying the number at risk over time. A vertical dashed line between 10 and 11 months indicates the point where the survival curves cross."
pl <- survfit2(Surv(month, evntd) ~ trt, data = ex6_crossing)
p <- pl |>
  ggsurvfit(theme = cowplot::theme_cowplot()) +
  add_quantile(x_value = 10.7, linetype = 2, color = pal_gray) +
  scale_x_continuous(
    breaks = seq(0, 27, 3),
    minor_breaks = seq(0, 27, 1)
  ) +
  add_risktable(risktable_stats = "n.risk") +
  theme(legend.position = "bottom") +
  scale_color_discrete(labels = c("Control", "Experimental")) +
  guides(x = guide_axis(minor.ticks = TRUE))
p
```

## Smoothed hazard rate estimates

We apply the bshazard package to smooth the hazard rate over time within
each treatment group.
Here, we used the default `nk = 31` knots for smoothing.
The [`bshazard()` documentation](https://rdrr.io/cran/bshazard/man/bshazard.html)
suggests a maximum of `nk = 40`, but higher values can be used
to see behavior induced by the `nk` parameter.
For instance, increasing to `nk = 120` likely will give a much less smooth curve.
Thus, selecting this parameter can be a bit of an art, given the
descriptive nature of the analysis.

```{r, results="hide"}
fit <- bshazard(
  Surv(month, evntd) ~ 1,
  data = subset(ex6_crossing, trt == 0),
  nk = 31
)
fit2 <- bshazard(
  Surv(month, evntd) ~ 1,
  data = subset(ex6_crossing, trt == 1),
  nk = 31
)
```

We plot the hazard rates with confidence bands by treatment group.

```{r}
#| smoothed-hazard-rates,
#| warning=FALSE,
#| fig.height = 4.6 / 1.2,
#| fig.cap="Smoothed hazard rate estimates with 95% confidence intervals for Control and Experimental groups over time. Dashed lines mark the crossing point before 6 months, highlighting where the hazard rates of the two groups intersect."
haz <- bind_rows(
  data.frame(
    Month = fit$time, Hazard = fit$hazard,
    Treatment = "Control",
    lower.ci = fit$lower.ci, upper.ci = fit$upper.ci
  ),
  data.frame(
    Month = fit2$time, Hazard = fit2$hazard,
    Treatment = "Experimental",
    lower.ci = fit2$lower.ci, upper.ci = fit2$upper.ci
  )
)

ggplot(haz, aes(Month)) +
  geom_line(aes(y = Hazard, col = Treatment)) +
  geom_ribbon(
    aes(ymin = lower.ci, ymax = upper.ci, fill = Treatment),
    alpha = 0.2, linetype = 0
  ) +
  scale_x_continuous(
    breaks = seq(0, 27, 3),
    minor_breaks = seq(0, 27, 1)
  ) +
  geom_segment(
    aes(x = 5.55, y = 0, xend = 5.55, yend = 0.065),
    linetype = 2, color = pal_gray, linewidth = 0.5
  ) +
  geom_segment(
    aes(x = 0, y = 0.065, xend = 5.55, yend = 0.065),
    linetype = 2, color = pal_gray, linewidth = 0.5
  ) +
  ylab("Hazard rate") +
  ylim(0, 0.1) +
  theme(legend.position = "bottom") +
  guides(x = guide_axis(minor.ticks = TRUE))
```

### Compare smoothed to actual survival function

While bshazard has been published [@rebora2014bshazard],
it is relatively new, and it is worth checking the model fit.
For this purpose, we transform the smoothed hazard rate estimates into
survival curves and compare them to the Kaplan-Meier curves previously estimated.
The Kaplan-Meier curves are reproduced accurately by the smoothed estimate
for each treatment group.
We noted previously that the Kaplan-Meier curves crossed after 10 months
for the treatment groups.
For hazard rates, we see the crossing of observed curves is before 6 months,
suggesting a different interpretation of when an excess risk period ends
for the experimental group compared to the control.

```{r}
#| smoothed-vs-km-survival-estimates,
#| fig.cap="Comparison of smoothed survival estimates and Kaplan-Meier survival curves. The smoothed survival curves (dashed lines) closely match the Kaplan-Meier estimates (solid lines), validating the smoothing approach."
# Get a smoothed survival estimate
pl2 <- haz |>
  group_by(Treatment) |>
  mutate(
    CHaz = cumsum((Month - lag(Month, default = 0)) * Hazard),
    Survival = exp(-CHaz), Estimate = "Smoothed"
  ) |>
  select(Survival, Month, Treatment, Estimate)

# Get Kaplan-Meier data from plots
km <- p$data |>
  as.data.frame() |>
  transmute(
    Survival = estimate,
    Month = time,
    Treatment = ifelse(strata == 1, "Experimental", "Control"),
    Estimate = "Kaplan-Meier"
  )

# Combine and plot
ggplot(
  bind_rows(pl2, km),
  aes(x = Month, y = Survival, col = Treatment, lty = Estimate)
) +
  geom_line() +
  scale_x_continuous(
    breaks = seq(0, 27, 3),
    minor_breaks = seq(0, 27, 1)
  ) +
  ylab("Survival") +
  ggtitle("Smoothed vs. Kaplan-Meier Survival Estimates") +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 11)
  ) +
  guides(
    color = guide_legend(order = 1),
    linetype = guide_legend(order = 2),
    x = guide_axis(minor.ticks = TRUE)
  )
```

The following example draws the Kaplan-Meier curve as a step function.

```{r}
#| km-step-functions-with-smoothed-survival,
#| fig.cap="Kaplan-Meier survival curves displayed as step functions (solid lines) with smoothed survival estimates (dashed lines)."
pl |>
  ggsurvfit(theme = cowplot::theme_cowplot()) +
  scale_x_continuous(
    breaks = seq(0, 27, 3),
    minor_breaks = seq(0, 27, 1)
  ) +
  geom_line(
    data = pl2 |> filter(Treatment == "Experimental"),
    aes(x = Month, y = Survival),
    col = pal_expt, linetype = "dashed", inherit.aes = FALSE
  ) +
  geom_line(
    data = pl2 |> filter(Treatment == "Control"),
    aes(x = Month, y = Survival),
    col = pal_ctrl, linetype = "dashed", inherit.aes = FALSE
  ) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 11)
  ) +
  scale_color_discrete(labels = c("Control", "Experimental")) +
  ggtitle("Smoothed vs. Kaplan-Meier Survival Estimates") +
  xlab("Month") +
  ylab("Survival") +
  # Add a legend for linetype
  geom_line(
    data = data.frame(
      x = c(0, 0, 0, 0), y = c(0, 0, 0, 0),
      Estimate = factor(rep(c("Kaplan-Meier", "Smoothed"), 2))
    ),
    aes(x = x, y = y, linetype = Estimate), inherit.aes = FALSE
  ) +
  guides(
    color = guide_legend(title = "Treatment", order = 1),
    linetype = guide_legend(title = "Estimate", order = 2),
    x = guide_axis(minor.ticks = TRUE)
  )
```

## Session info

```{r, echo=FALSE}
pkgs <- sessioninfo::session_info(pkgs = "attached", info = "packages")
pkgs$packages$library <- NULL
pkgs
```

## References
